\MakeUppercase{Linear versus centred chromatic numbers}\thanks{This research was partly funded by NSERC.}

### 2024-04-04 referee-response.md

The referee had 4 broad comments:

**Referee Comment 1:** The paper's title is a bit ambitious since, as the authors point out, the machinery for relating the linear chromatic number to the centred one was already introduced previously (and this manuscript does not develop on it at all). A more honest title would be something along the lines of "The linear chromatic number of pseudogrids".

**Response:** We have chosen to leave the title as is.  In our opinion, the current title doesn't promise much.  The proposed title, on the other hand, suggests that the results apply only to pseudogrids.  Consider, for example, the similar-sounding title "The linear chromatic number of cactus graphs."

With the current title a reader will still learn, within the first two pages that the improvement comes only by studying the linear chromatic number pseudogrids and decide for themselves whether it's worthwhile reading the rest of the argument.  (These first two pages also have the benefit of catching-up a reader on the current state of the art and the limitations of Kun et al's argument.)

**Referee Comment 2:** There are a few places (e.g. in section 3.6 -- see PDF) where the authors cut a few corners in their proofs; this should be addressed, but I don't think it is anything serious.

@@

We have gone through the attached PDF carefully and corrected the minor issues pointed out. Some larger changes in response to the referee's comments are discussed below.

**Referee Comment 3:** A lot of the arguments in the middle sections, such as the packing lemma, seem to use the underlying intuition that grids (and by extension pseudogrids) look like the plane with the taxicab metric and the Lebesgue measure. I imagine the authors have already attempted this, but is there really no way to make that intuition rigorous by defining a metric and/or a measure on appropriate sets, and express the lemmas in terms of those? Something like that would make arguments such as the one in Lemma 13 much cleaner. In this direction, have the authors considered the idea of using a grid with every edge subdivided once as the index set for the path partition of the pseudogrid (the subdivision vertices symbolising the edges), with the purpose of making definitions such as the r-boxes more uniform (and of having a symmetric relation, whereby if \mu is in the r-box of \nu, then \nu should be in the r-box of \mu)?

During the initial writing phase of this result, we tried several such strategies.  The main annoying issue is that mapping the vertices of a subdivided grid onto objects in the pseudogrid sometimes maps a vertex onto a single edge of the pseudogrid (but not the endpoints of that edge).  In terms of colouring, this means that the subdivision vertex has an empty colour set.  It is possible to define things carefully enough for this to work, but then the reader is constantly having to verify that simple-sounding statements are correct under these careful definitions.

**Referee Comment 4:** In the acknowledgements, the authors mention that it was pointed out to them the problem on a grid (as opposed to pseudogrid) has a simple solution. Assuming the person who pointed that out is happy with it, it might be interesting to include the solution.

@@

TODO

---

**Referee Comment:** Bottom of Page 2: A tiny bit misleading (since as far as I can tell, the bottleneck for getting conjecture 2 is the grid theorem, not this).

**Referee Comment:** Top of Page 3: This somewhat addresses the previous comment.

**Response:** We have added further explanation (at the location of the second comment) of the limits of this argument here. In particular, even with an Excluded Grid Theorem that matches the currently known lower bound, this would only establish that the centred chromatic number is at most cubic in the linear chromatic number.  Even for planar graphs, the argument can only show that the centred chromatic number is at most quadratic in the linear chromatic number.

@@

**Referee Comment:** I do not understand how this works when removing edges in $P_{v_i}$ -- doesn't $P_{v_i}$ belong to both a row and a column (and thus deleting it leaves us with both fewer rows and columns?)

**Response:** The referee is correct.  The description of deletion has been corrected.  Now we only delete edges of $G$ that correspond to edges of the grid (and then repeatedly remove vertices of degree at most $1$).

**Referee Comment:** $d$ has not been defined. The last part of the proof (where it is shown that a certain value of $d$ would work) should come first -- or at the very least, $d$ should be chosen bigger than some bound given here (and the last part of the proof would show this bound works). This is important, since I am having some trouble keeping track of the dependencies between the variables, so that should be made more transparent.

**Response:** The variable $d$ is defined in the statement of the lemma ("there exists a value $d\in O(r^4\log r)$ such that....").  In the first line of the proof, we use this value of $d$ to apply Lemma 9 to obtain an assignment of colours of objects so that each colour is assigned to at least $d$ objects.  We have expanded the first line of the proof to point this out, so that it sounds familiar when it is mentioned again at this point in the proof.

@@

**Referee Comment:** From what I can tell, we are saying that if there are enough elements of $\phi^{-1}(\alpha)$ far from the elements of $Q_1$, there will be two such elements far from each other. I agree with this in principle, but where does the actual bound given here come from? The intuition, I presume, is that a set of that size is "maximal" of diameter at most $2r + 1$, but how is that made rigorous? Presumably some argument involving convexity,  and it would be good to see it written.

**Response:** It is much simpler than this.  During Round 1, when an object is added to $Q_1$, it *covers* all the nearby objects.  Round 1 succeeds for $\alpha$ if, when the colour $\alpha$ is considered, there exists two uncovered vertices $\mu_1$ and $\mu_2$ of colour $\alpha$ that are "far" from each other.  For a single object $\mu_1$, the number of vertices that "close" to $\mu_1$ is $\boxplus_r$ (by definition).   So if there are $\boxplus_{r+1}>\boxplus_r$ uncovered objects of colour $\alpha$, we can choose any of them to be $\mu_1$, eliminate the at most $\boxplus_{r}$ objects that are close to $\mu_1$, and then take $\mu_2$ to be any of the remaining objects of colour $\alpha$.  We have added a paragraph describing this argument.

### 2024-04-04 lin-vs-cen.tex
More changes suggested by the referee.

  We now want to use the Local Lemma (\cref{weighted_lovasz}) to show that, with probability greater than zero, $\deg_H(\alpha)\ge 2\vol{2r+1}$ for each $\alpha\in X$.  For each $\alpha\in X$, consider the set $\phi^{-1}(\alpha)\subseteq \VE(G_k)$ and recall that, since $\phi$ comes from \cref{one_colour_per_object}, $|\phi^{-1}(\alpha)| \ge d$.  Furthermore all but at most $\vol{r+1}$ elements of $\phi^{-1}(\alpha)$ are contained in $\bigcup_{\mu\in Y}\VE(G_{2r+1}(\mu))$ since otherwise the first round would have been an $\alpha$-success.  Let $\Phi_\alpha:=\phi^{-1}(\alpha)\cap \left(\bigcup_{\mu\in Y}\VE(G_{2r+1}(\mu))\right)$, so $|\Phi_\alpha|\ge d-\vol{r+1}$.  In words, $\Phi_\alpha$ consists of all objects of colour $\alpha$ that were eliminated from consideration because they are too close to an element of $Y$.

@@

  We begin by showing that, for each $\alpha\in X$, the random variable $\deg_H(\alpha)$ dominates\footnote{A random variable $X$ dominates a random variable $Y$ if $\Pr(X\ge x)\ge \Pr(Y\ge x)$ for all $x\in\R$.} a $\binomial(\ceil{|\Phi_\alpha|/\vol{7r}},1/16)$ random variable, which allows us to establish an exponential inequality for $\Pr(\deg_H(\alpha) < 2\vol{2r+1})$. Suppose that, for some $\alpha\in X$ and $\mu\in Y$, $B_{2r+1}(\mu)\cap \Phi_\alpha$ contains at least one element $\nu$ and consider the set
  \[
     Y_\nu := \{\xi\in Y:\nu\in\VE(G_{3r+1}(\xi))\} \enspace .
  \]
  The edge $\alpha\mu$ will appear in $H$ if, in our random permutation $\pi$ of $Y$, $\nu$ appears before any other element of $Y_\nu$.  By \cref{packing_lemma}, $|Y_\nu|\le 16$, so the probability that the edge $\alpha\mu$ appears in $H$ is at least $1/|Y_\nu|\ge 1/16$. This is already enough to establish that $\E(\deg_H(\alpha)) \in \Omega(|\Phi_\alpha|/r^2)$, but in order to obtain a sufficiently strong concentration result for $\deg_H(\alpha)$ we need to find some independence.\footnote{There are many ways to obtain a concentration result here and we use a simple method with the goal of being self-contained.  For a tighter result, use the method of bounded differences \cite{mcdiarmid:on}.}  To do this, set $J:=\Phi_\alpha$ and consider the greedy process of repeatedly choosing some $\mu\in Y$ such that $\VE(G_{2r+1}(\mu))\cap J$ contains at least one element $\nu$ and then setting $J:=J\setminus \VE(G_{7r}(\nu))$.\footnote{The constant $7$ is overkill here and is only used for simplicity; $6r+O(1)$ would be sufficient.}  This process continues until $J$ is empty. Since $|\VE(G_{7r}(\mu))|\le \vol{7r}$, each iteration in this process eliminates at most $\vol{7r}$ elements from $J$ so the number, $t$, of iterations is at least $\ceil{|\Phi_\alpha|/\vol{7r}}$.

@@

  Comparing \cref{weight_product,probability} we see that both quantities decrease exponentially in $|\Phi_\alpha|$ but \cref{weight_product} does so at a rate that can be controlled with $\tau$.
  Taking $\tau = {32\vol{7r}^2}$, and using \cref{weight_product,probability}, we find that
  \[
     \Pr\left(E_\alpha\right) \le w(E_\alpha)\prod_{\beta\in\Gamma_\alpha}(1-w(E_\beta))
  \]
  provided that
  \[   |\Phi_\alpha| \ge {32\vol{7r}}\big(\ln({32\vol{7r}}+1) + {2\vol{2r+1}}\ln(2|\Phi_\alpha|)\big) \enspace .
  \]
  The right hand side of this last equation is of the form $O(r^2\log r+r^4\log|\Phi_\alpha|)$.  Since $|\Phi_\alpha|\ge d-\vol{r+1}\in d- O(r^2)$, this can therefore be satisfied for some $d\in O(r^4\log r)$.

@@

\begin{proof}[Proof of \cref{pseudogrid_lower_bound}]
  Suppose for the sake of contradiction that $G$ has a linear colouring $\varphi$ using fewer than $\epsilon k$ colours for arbitrarily small $\epsilon >0$.  Then, by \cref{only_frequent}, $G$ contains a $k'\times k'$ pseudogrid $G'$ with $k'\ge k/2$ having a grid-decomposition $\mathcal{P}':=\{P_\mu':\mu\in \VE(G_{k'})\}$ with the property that, for any $A\subseteq\varphi(V(G'))$, $|\varphi^{-1}_{\mathcal{P'}}(A)\cap\interior_r(G_{k'})|\ge d|A|$, for any $d\le 1/2\epsilon-2$.  In particular, for some $\epsilon \in \Omega(1/r^4\log r)$, \cref{doubled_colour_set} implies that $G'$ contains a set $S$ of $2|\varphi(V(G')|$ vertices containing two vertices of each colour in $\varphi(V(G'))$ such that $|B_r(\mu)\cap S|\le 2$ for each $\mu\in S$.  By \cref{pick_up_everything}, $G'$ contains a path $P$ that contains every vertex in $S$.  Since $P$ is contained in $G'$, $\varphi(V(P))=\varphi(V(G'))$, so $P$ is has no center under $\varphi$, contradicting the assumption that $\varphi$ is a linear colouring of $G$.
\end{proof}

>>>>

  We now want to use the Local Lemma (\cref{weighted_lovasz}) to show that, with probability greater than zero, $\deg_H(\alpha)\ge 2\vol{2r+1}$ for each $\alpha\in X$.  For each $\alpha\in X$, consider the set $\phi^{-1}(\alpha)\subseteq \VE(G_k)$ and recall that, since $\phi$ comes from \cref{one_colour_per_object}, $|\phi^{-1}(\alpha)| \ge d$.

  For each $\alpha\in X$, let $\Phi_\alpha:=\phi^{-1}(\alpha)\cap \left(\bigcup_{\mu\in Y}\VE(G_{2r+1}(\mu))\right)$.  In words, $\Phi_\alpha$ consists of all objects of colour $\alpha$ that were eliminated from consideration because they are too close to an element of $Y$. We claim that, for each $\alpha\in X$, $|\Phi_\alpha|\ge d-\vol{r}$.  Otherwise $|\phi^{-1}(\alpha)\setminus\Phi_\alpha| > \vol{r}$. Then, when the colour $\alpha$ was considered during the first round, we could have taken $\mu_1$ to be any object in $\phi^{-1}(\alpha)\setminus\Phi_\alpha$ and taken $\mu_2$ to be any object in $\phi^{-1}(\alpha)\setminus\Phi_\alpha\setminus B_{2r+1}(\mu_1)$. (Recall that $\vol{r}\ge |B_{2r+1}(\mu_1)|$, by definition.)  Thus, the first round would have been an $\alpha$-success, contradicting the fact that $\alpha\in X$.

@@

  We now show that, for each $\alpha\in X$, the random variable $\deg_H(\alpha)$ dominates\footnote{A random variable $X$ dominates a random variable $Y$ if $\Pr(X\ge x)\ge \Pr(Y\ge x)$ for all $x\in\R$.} a $\binomial(\ceil{|\Phi_\alpha|/\vol{7r}},1/16)$ random variable, which allows us to establish an exponential inequality for $\Pr(\deg_H(\alpha) < 2\vol{2r+1})$. Suppose that, for some $\alpha\in X$ and $\mu\in Y$, $B_{2r+1}(\mu)\cap \Phi_\alpha$ contains at least one element $\nu$ and consider the set
  \[
     Y_\nu := \{\xi\in Y:\nu\in\VE(G_{3r+1}(\xi))\} \enspace .
  \]
  The edge $\alpha\mu$ will appear in $H$ if, in our random permutation $\pi$ of $Y$, $\nu$ appears before any other element of $Y_\nu$.  By \cref{packing_lemma}, $|Y_\nu|\le 16$, so the probability that the edge $\alpha\mu$ appears in $H$ is at least $1/|Y_\nu|\ge 1/16$. This is already enough to establish that $\E(\deg_H(\alpha)) \in \Omega(|\Phi_\alpha|/r^2)$, but in order to obtain a sufficiently strong concentration result for $\deg_H(\alpha)$ we need to find some independence.\footnote{There are many ways to obtain a concentration result here and we use a simple method with the goal of being self-contained.  For a tighter result, use the method of bounded differences \cite{mcdiarmid:on}.}  To do this, set $J:=\Phi_\alpha$ and consider the greedy process of repeatedly choosing some $\mu\in Y$ such that $\VE(G_{2r+1}(\mu))\cap J$ contains at least one element $\nu$ and then setting $J:=J\setminus \VE(G_{7r}(\nu))$.\footnote{The constant $7$ is overkill here and is only used for simplicity; $6r+O(1)$ would be sufficient.}  This process continues until $J$ is empty. Since $|\VE(G_{7r}(\mu))|\le \vol{7r}$, each iteration in this process eliminates at most $\vol{7r}$ elements from $J$ so the number, $t$, of iterations is at least $\ceil{|\Phi_\alpha|/\vol{7r}}$.

@@

  Comparing \cref{weight_product,probability} we see that both quantities decrease exponentially in $|\Phi_\alpha|$ but \cref{weight_product} does so at a rate that can be controlled with $\tau$.
  Taking $\tau = {32\vol{7r}^2}$, and using \cref{weight_product,probability}, we find that
  \[
     \Pr\left(E_\alpha\right) \le w(E_\alpha)\prod_{\beta\in\Gamma_\alpha}(1-w(E_\beta))
  \]
  provided that
  \[   |\Phi_\alpha| \ge {32\vol{7r}}\big(\ln({32\vol{7r}}+1) + {2\vol{2r+1}}\ln(2|\Phi_\alpha|)\big) \enspace .
  \]
  The right hand side of this last equation is of the form $O(r^2\log r+r^4\log|\Phi_\alpha|)$.  Since $|\Phi_\alpha|\ge d-\vol{r}\in d- O(r^2)$, this can therefore be satisfied for some $d\in O(r^4\log r)$.

@@

\begin{proof}[Proof of \cref{pseudogrid_lower_bound}]
  Suppose, for the sake of contradiction, that $G$ has a linear colouring $\varphi$ using fewer than $\epsilon k$ colours where $\epsilon:=1/(d+2r)$, $r:= 13$ and $d\in O(r^4\log r)=O(1)$ is some large integer constant.  Then, by \cref{only_frequent}, $G$ contains a $k'\times k'$ pseudogrid $G'$ with $k'\ge k/2$ having a grid-decomposition $\mathcal{P}':=\{P_\mu':\mu\in \VE(G_{k'})\}$ with the property that, for any $A\subseteq\varphi(V(G'))$, $|\varphi^{-1}_{\mathcal{P'}}(A)\cap\interior_r(G_{k'})|\ge d|A|$.  \Cref{doubled_colour_set} implies that $G'$ contains a set $S$ of $2|\varphi(V(G')|$ vertices containing two vertices of each colour in $\varphi(V(G'))$ such that $|B_r(\mu)\cap S|\le 2$ for each $\mu\in S$.  By \cref{pick_up_everything}, $G'$ contains a path $P$ that contains every vertex in $S$.  Since $P$ is contained in $G'$, $\varphi(V(P))=\varphi(V(G'))$, so $P$ is has no center under $\varphi$, contradicting the assumption that $\varphi$ is a linear colouring of $G$.
\end{proof}
